\hydrogen{} and \carbon{} Network Strategies for the European Energy System

### 2024-09-09 report/submissions/journal/review-responses/review-responses-first-round.md

We thank both reviewers for their helpful comments. We have rerun all scenarios in the paper with improved assumptions as suggested by the authors (improved accounting for leakage and compression losses) as well as new scenarios requested by the reviewers for a sensitivity to sequestration rates.

## Reviewer #1 (Remarks to the Author):

@FabianHofmann
> The article presents an interesting analysis of the foreseen H2 and CO2 transport infrastructures.
> Literature is not accurate and there are many references cited without going in their details.
> The study lacks of realistic elements such as leakage issues, location of compression stations in relation with the capture location.

Thank you for your comments. We appreciate your feedback and will address your concerns in the revised manuscript.

We have expanded the literature review section to provide more context for the study, as well as discussing recent policy measures for carbon management in the EU. Since the study touches a wide range of topics in the political and technological landscape, we have tried to find a good balance between providing a comprehensive overview and keeping the text concise. (#TODO)

To simulate a more realistic transport system, the following assumptions have been integrated into the energy system model for transport efficiencies and compression requirements. For hydrogen pipelines an average loss of 1.2% per 1000 km, aligning with industry estimates that account for hydrogen's high permeability and small molecular size, which can lead to higher losses compared to other gases​. The compression energy required is assumed to be 1.8% of the transported energy, which is consistent with the significant energy demands for compressing hydrogen due to its low energy density​ [Environmental Defense Fund​](https://www.edf.org/sites/default/files/documents/Pipeline%20Methane%20Leaks%20Report.pdf). For natural gas pipelines, the model assumes a 1% loss rate per 1000 km, accounting for typical leakage and operational losses in well-maintained natural gas infrastructure​ as reported in [Global Energy Monitor](https://www.gem.wiki/Natural_gas_transmission_leakage_rates). The compression requirement is set at 1%, which falls within the common range for natural gas pipeline operations​ [Environmental Defense Fund​](https://www.edf.org/sites/default/files/documents/Pipeline%20Methane%20Leaks%20Report.pdf)​. For CO2 pipelines an average loss rate of 0.8% per 1000 km is reflecting a generally low leakage rate observed in supercritical or dense phase CO2 transport​ [SpringerLink](https://link.springer.com/article/10.1007/s11356-023-27289-3) [MDPI](https://www.mdpi.com/1996-1073/14/15/4601)​. The compression energy needed is set to 250 kWh per 1000km and tonne CO2 reflecting the energy required to compress and maintain CO2 in a supercritical state for efficient transport​.

---
@FabianHofmann
> There are several graphs that make the reader confused rather than explain the amount of work done by the authors.

We have revised the layout of the figures to make them more accessible to the reader and have adjusted the color scheme where colors tended to be quite similar. However, we would appreciate more specific feedback on which figures you find confusing so that we can address these issues more effectively. Most of the figures contain a lot of information that we feel is necessary to support the argument. We have tried to make them as clear as possible.

---

@nworbmot
> The results are not understandable in terms of transferring into policies.

We have written a short policy outlook in the conclusion as well as linking in the introduction to the recently-announced industrial carbon management strategy of the European Commission. Our addition to the conclusion reads:

"For policy makers these results show the need for coordinated planning across sectors like carbon, hydrogen and synthetic fuels since they are strongly interacting, as well as across borders because of the localized nature of good renewable resources and sequestration potentials. Deploying multiple networks offers some robustness should unforeseen problems arise with one of them, and our results show that the system can even cope with neither a CO2 nor H2 network for a cost penalty.  Many of the technologies have not yet been deployed at scale before and will need support for financing, regulation and gaining the acceptance of local populations."

---

## Reviewer #2 (Remarks to the Author):
@FabianHofmann

> Overall, a very impressive paper, with a very clear and concise structure, easy to follow with many interesting insights. The research gap and necessity of the paper clearly are demonstrated, the topic is higly relevant and the chosen methodology are well suited for the task.

> There are a couple of aspects I would like to understand better, i.e. aspects in which I think additional information would be beneficial.

> 1: CCS is a challenge for scenario analyses like yours, since it low cost makes it a backstop technology. The model will try to solve many things with CCS if one does not prevent somehow. Your approach is to limit CS. “This is in order to avoid offsetting fossil emissions with carbon removal if they can technologically be avoided in the first place.” I was wondering about this throughout the paper. So, first of all I could be helpful to point out the importance and weight of this setting early in the text. I think the implications of this should be discussed more, because I believe this to be crucial: At your point of operation, your model is able to provide and sequester CO2 for roughly160 €/t if I’m correct. That basically means that all mitigation measures costlier than that are only part of the solution because of the CS limit. It would be interesting to discuss the dual value/shadow price of tis cap, to understand how much economic pressure could be on relying more on CCS. The fossil lock-in your CS cap prevents could be an economically very attractive. So, if compensating fossil fuel usage is cheaper than the chosen path of FT and other PtX options, I think this would deserve a discussion in the paper. Especially because the compensation would not need to take place in Europe but could be done in regions with even lower DACS/BECCS costs.

Thank you for your insightful comment. In response, we have added an explanation for our choice to the introduction and have added new modelling runs in a new subchapter in the Appendix "Higher Sequestration Limits" which we refer to in the main text. The added chapter discusses new model results for higher seqestration limits and their implications.

Our analysis shows that increasing the limit from 200 Mt/year to 800 Mt/year reduces system costs quite significantly, by up to 9.1%. This reduction is primarily due to a shift towards greater reliance on fossil fuels either with CCS or unabated and compensated with CDR, which in turn reduces the need for CU and renewable investments by about a third. At the same time, the hydrogen network loses importance whereas the carbon network quadruples its capacity. Point-sources of carbon in the inland where fossil fuels are used, like gas for heating or naphtha for industry, are nearly all captured and tranpsorted to be sequesterd.

Another interesting finding is the drop of the shadow price from €225/t to €51/t when increasing the sequestration limit. Additional sequestration capacity provides economic benefits, these benefits, however, diminish as the limit increases​.

However, this also introduces the risk of reinforcing fossil fuel dependence, potentially undermining long-term decarbonization efforts. The discussion in the appendix now highlights these trade-offs, emphasizing the importance of carefully balancing sequestration limits to avoid economic over-reliance on CCS​.

---

@chrstphtrs
> 2. The H2 provision in the Baseline case confuses me a bit, namely that it works surprisingly well. How is a constant supply possible in regions without geological storage potential? Does this mean that in hours of low/no RES power plants in the regions with storages produce electricity, which is then brought to H2 sinks and turned into H2 again? In our scenarios this led to much bigger costs and almost impossible necessities for electricity network expansions. Is this because your H2 demand is relatively low, and you transform it into other carriers? Or do you have some other form of storage option at the consumption sites. Overall, I would just like to understand better how you supply H2 steadily without a network.

Thank you for the comment, this is indeed a very interesting question.

The spatially-fixed hydrogen demand from industry (primarily for primary steel but also for ammonia) is comparatively low, while the bulk of the H2 demand (FT, methanolisation, methanation) can be moved by the model to locations with low-cost renewables and geological H2 storage. This avoids over-burdening the electricity grid.

In addition, to provide hydrogen in some regions with poor renewable resources the model resorts to SMR with either green or fossil methane transported in the gas grid.

---
@fneum
> 3: In the introduction: “It is assumed that Europe is self-sufficient in energy and does not import any fuels.” I would like to see the consequences of this setting discussed, especially regarding the large liquid quantities it entails. Just a suggestion, but a variation in which half of the quantities are imported discussed briefliy in the Annex would be very valuable. Much of the infrastructure you model and discuss discuss hinges on this, and importing a major share is clearly an realistic and attractive option in reality.

This is indeed an important point that we have not discussed in detail in the paper. The assumption of Europe's self-sufficiency in energy was made to simplify the system design and focus on internal infrastructure. It also has the advantage of making good use of Europe's sustainable biomass sources for sourcing CO2, and avoiding the difficulty of regulating sustainable CO2 provision for carbonaceous fuels synthesised outside of Europe. However, we agree that importing a significant share of energy carriers, particularly liquid fuels, is a realistic and potentially cost-beneficial option.

We have expanded the discussion of the limitations in the appendix to explore the potential consequences of allowing imports. In the higher sequestration scenario, for example, our model shows that increased imports of fossil oil significantly reduce the need for domestic Fischer-Tropsch (FT) fuel production. While this reduces hydrogen transport and carbon utilization requirements in Central Europe, it also necessitates substantial carbon sequestration within Europe to offset emissions from fossil-based imports. If these imports were carbon-neutral, the need for dedicated sequestration infrastructure would be significantly reduced, as the emissions would already be compensated.
Thus, the carbon neutrality of imported fuels has a profound effect on the optimal infrastructure layout, especially for hydrogen transport and carbon sequestration. A scenario that allows for partial energy imports could lead to lower investments in the carbon and hydrogen networks as well as direct air capture.

However, a critical political and strategic consideration is the control over the CO2 supply chain. There may not be sufficient sustainable CO2 available abroad, and Europe may prefer to maintain control of the CO2 supply chain to ensure security and reliability. Relying on external sources for carbon-neutral fuels could introduce geopolitical risks, similar to those faced with fossil fuel imports.

---
@chrstphtrs
> Your H2 prices seem comparatively low to similar exercises. What interest rate/ WACC did you use? I think that should always be made explicit, as it has huge a impact on many results.

We use technology cost assumptions from the Danish Energy Agency referenced in the Methods section. The database applies a discount rate of 7% to derive annualized investment costs for all major technologies, including hydrogen electrolyzers. We have added this information to the Methods section.

---
@FabianHofmann
> Maybe this would be to much, but while seeing your H2 prices I was very interested in seeing your electricity prices. If a map/maps would be too much, maybe state the ranges somewhere for reference?

We have added a chapter to the Appendix that showes maps for other energy carrier (electricity, heat, gas, biomass) in the hybrid scenario. These show the average production, consumption, transport and prices per carrier. We hope this will provide the information you are looking for.

---
@chrstphtrs
> I think an overview table with differences of the scenario settings in the beginning would be helpful, and maybe with some key results in the end.

We have added an overview table in the introduction that shows the main differences between the scenarios. Since the difference of the scenarios is important on the qualitative level, we have not added a table with key results in the end. We hope that the reader can follow the argumentation and the results are clear enough.

---
@FabianHofmann
> The colour coding in Figure 6 could be better, the greens are almost indistinguishable and too many similar browns.

Thank you for your feedback. We have revised the color scheme in Figure 6 to make it more accessible to the reader. We hope that the new color scheme will make the figure easier to read.

---
@chrstphtrs
> In the end of section 2: “In the network layouts, the slight shift in CU deployment results in a partial dismantling of hydrogen pipelines, especially from the Iberian Peninsula to Central Europe (0.5 bn€/a).” Is “dismantling” really the right word here? The model simply does not build those pipelines in this scenario, and you cannot dismantle what you never built. Or maybe you can, and this is a tricky koan.

You are absolutely correct. We rephrased this passage, and similar ones, that the the net-zero and net-negative scenarios are modelled as different futures, not as chronological events.

---
@chrstphtrs
> “High-cost Direct Air Capture”: I may be nitpicking, but I was wondering whether this expression is really fair in combination with the CS cap. I think you force the model to pursue many even more expensive option like FT, this is just the most expensive carbon option for sequestration. Again, really just a thought.

You are correct, the relative statement of low and high cost only applies to elementary carbon dioxide needed for sequestration. We amended the statement to make it more clear.

### 2024-09-09 report/submissions/journal/paper/paper.tex
address internal review

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an essential role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. For the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore both networks' potential synergies and competition in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from biomass and reducing reliance on direct air capture. Our analysis suggests integrating hydrogen and carbon dioxide networks into European energy policy for a robust, carbon-neutral or carbon-negative future energy system.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\carbon{}) and hydrogen (\hydrogen{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is being considered in several industries that are not eligible for electrification. It offers an efficient way to transport and store energy over long distances and periods and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations to extract it from the atmosphere permanently, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels, such as synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstock. We jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \carbon{} and \hydrogen{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

@@

At the same time, collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier.
% Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}.
Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer to purchase \carbon{} on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning with a transport and sequestration capacity of 1.5~Mt \carbon{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential~\cite{ToolsGreenTransition} of 1.7 Gt carbon annually from point sources, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it is important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmannNETZEROGEOSPHERICRETURN}.

Up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steel production, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several publications, all of which, however, dealt with the isolated aspects~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to Integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks, which are crucial to provide a holistic view of the energy system and its technological interactions. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the interchangeability of hydrogen and electricity grid expansion in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses the most relevant sectors, however it neglects the option to transport carbon across regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects sectors like industry that will need to handle most \carbon{} in the future.
The European Commission's Joint Research Center analysed the possible topology of a \carbon{} network in 2024~\cite{jrc2024}, but didn't include the interaction of  \carbon{} sources and sinks with the rest of the energy system.
A comprehensive example is found in~\cite{becattiniCarbonDioxideCapture2022}, which presents a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connected to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and the co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. Euro-Calliope, for example, does not feature carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on captive electrolysis and DAC in the same region~\cite{pickeringDiversityOptionsEliminate2022}.

No study has yet considered the co-optimization and comprehensive assessment of both \carbon{} and \hydrogen{} networks in a fully sector-coupled energy system. However, such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model~\cite{brownPyPSAEurOpenSectorcoupled2023} which optimizes operations and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuel production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

@@

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. Initially, we focus on a net-zero carbon emissions target, limiting carbon sequestration to 200~Mt/a. This constraint ensures that carbon removal is reserved for the hardest-to-abate emissions, while avoiding reliance on sequestration where other mitigation options are feasible. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net \carbon{} removal target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 764~bn€a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.7\% in the \hydrogengrid{}, and 5.3\% in the \hybridscenario{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost, subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While in the \baselinescenario{}, the model has neither a carbon nor a hydrogen network, it can expand both in the \hybridscenario{}. ``Gas Infrastructure'' combines gas facilities for power and heat production, ``\carbon{} Infrastructure'' and ``\hydrogen{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP unit) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

System cost differences between the scenarios are driven by wind and solar, as well as carbon capture technologies. In scenarios with carbon and hydrogen networks, the model primarily reduces the reliance on high-cost DAC, from 365~Mt/a carbon removal in the \baselinescenario{} to 113~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas-to-gas upgrading increase, the latter lowering the need for synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

@@

In the \baselinescenario{}, where there is no carbon and hydrogen transport, hydrogen must be produced where it is consumed. For CU, either hydrogen must be produced where carbon is cheap at point sources, or DAC must be used at sites with low-cost hydrogen. Where carbon from point sources cannot be sequestered, it has to be compensated elsewhere by negative emissions from DAC or biomass capture with sequestration. All of these options are sub-optimal from a cost perspective.

@@

In the \carbonscenario, there are three main transport purposes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported to the Iberian Peninsula (9.8~kt/h) and the British Isles (11.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. In Italy and Greece, a portion of the carbon (3.3~kt/h) is directed to methanation facilities where the synthetic methane is fed into the gas grid at key terminals that also receive fossil gas imports. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point source. In addition to electrolysis in regions with good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8~GW$_\text{el}$) (see bottom left). Hydrogen prices vary widely across the continent, with low price ``valleys'' at around 60~€/MWh, where the CU is located, and high price ``peaks'' in Central Europe at around 95~€/MWh. Despite the transport system, carbon prices still vary from 105~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation route goes from Spain (66~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources (see Fig~\ref{fig:balance_map}, bottom right). Also, the hydrogen network supplies spatially fixed hydrogen demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration takes place at several coastal sites with local DAC facilities. The largest carbon sink is located in Portugal (3.5~kt/h, see Fig~\ref{fig:balance_map}, top right). Carbon prices are low in Central Europe (110-120~€/t) and highest in sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.

>>>>

\begin{abstract}
    Hydrogen and carbon dioxide transport can both play an essential role in climate-neutral energy systems. Hydrogen networks help serve regions with high energy demand, while excess emissions are transported away in carbon dioxide networks. For the synthesis of carbonaceous fuels, it is less clear which input should be transported: hydrogen to carbon point sources or carbon to low-cost hydrogen. We explore both networks' potential synergies and competition in a cost-optimal carbon-neutral European energy system. In direct comparison, a hydrogen network is more cost-effective than a carbon network, since it serves to transport hydrogen to demand and to point source of carbon for utilization. However, in a hybrid scenario where both networks are present, the carbon network effectively complements the hydrogen network, promoting carbon capture from distributed biomass and reducing reliance on direct air capture. The layouts of the hydrogen and carbon dioxide networks are robust if the climate target is tightened to be net negative.
\end{abstract}

@@

The transition to a climate-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both carbon dioxide (\carbon{}) and hydrogen (\hydrogen{}) is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including carbon capture, transport, storage, utilization, and sequestration. Hydrogen is being considered in several industries that are not eligible for electrification. It offers an efficient way to transport and store energy over long distances and periods and can be produced from electrolysis or steam methane reforming (SMR), for example. Captured carbon can be stored in geological formations permanently, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels, such as synthetic kerosene for aviation, synthetic methanol for shipping or synthetic methane for industrial feedstock or synthetic naphtha as a feedstock for high value chemicals. The production of these carbonaceous fuels is jointly referred to as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \carbon{} and \hydrogen{} networks are likely to play a crucial role in the cost-effective integration of these carbon management technologies to enable net-zero economies.

@@

At the same time, collaborative industry initiatives like the European Hydrogen Backbone~\cite{gasforclimateEuropeanHydrogenBackbone2022} or the Hydrogen Infrastructure Map~\cite{H2InfrastructureMap} showcase the potential of hydrogen as a fuel and energy carrier.
% Some natural gas pipelines have already been repurposed to transport hydrogen~\cite{RohrFreiFuer}.
Business models from companies like Tree Energy Solutions~\cite{TESHydrogenLife2023}, Carbfix~\cite{WeTurnCO2}, and Equinor~\cite{adomaitisEquinorRWEBuild2023} advertise carbon management hubs that provide green hydrogen, methane, and synthetic fuels on the one hand and offer to purchase \carbon{} on the other hand. The Northern Lights project in Norway~\cite{NorthernLightsWhat} is planning a transport and sequestration capacity of 1.5~Mt \carbon{} per year to be operative in 2024, expanding to a targeted scale of 5~Mt per year of sequestration by 2030.
The market potential for carbon capture and sequestration in Europe is underscored by the Capture Map's estimated potential of 1.7 Gt carbon annually from point sources~\cite{ToolsGreenTransition}, accounting for about half of the continent's emissions, as well as estimated sequestration capacity potentials of up to 3~Gt/a~\cite{europeancommissionEuropeanCO2Storage}. However, it is important to note that carbon sequestration technology, unlike carbon capture, is not yet fully mature and its full potential and implications remain to be explored in depth.
To advance the carbon economy, the Clean Air Task Force emphasizes the importance of developing a carbon transport system in Europe alongside a hydrogen network~\cite{lockwoodEuropeanStrategyCarbon}. Carbon pipelines, considered a mature technology, have seen widespread installations in the United States and Canada, primarily to supply \carbon{} to enhanced oil recovery~\cite{righettiSitingCarbonDioxide2017,friedmann2020net}.

Despite the abundance of policies and initiatives, up to this point, it remains unclear how the two transport systems of hydrogen and carbon may complement or even compete with each other. Both can bridge the misalignment of sources and sinks for carbon and hydrogen. A hydrogen network can supply regions with geographically fixed hydrogen demand, such as for steel production, with hydrogen from regions with the best renewable resources as well as enable CU at the site of point-source CC. On the other hand, the carbon transport system allows transporting captured carbon to regions with sequestration potentials or high-quality renewable resources, the latter enabling cost-effective CU.

In the literature, the two network approaches and underlying technologies have been discussed in several publications~\cite{bakkenLinearModelsOptimization2008,morbeeOptimisedDeploymentEuropean2012,stewartFeasibilityEuropeanwideIntegrated2014,oeiModelingCarbonCapture2014,elahiMultiperiodLeastCost2014,burandtDecarbonizingChinaEnergy2019,middletonSimCCSOpensourceTool2020,bjerketvedtOptimalDesignCost2020,weiProposedGlobalLayout2021,damoreOptimalDesignEuropean2021,becattiniCarbonDioxideCapture2022,neumannBenefitsHydrogenNetwork2022}. Such techno-economic models, in contrast for example to integrated Assessment Models, can account for the spatial distribution of carbon sources and sinks, which are crucial to provide a holistic view of the energy system and its technological interactions. However, all of the techno-economic studies have only dealt with isolated aspects of hydrogen and carbon networks. Neumann et al.~\cite{neumannBenefitsHydrogenNetwork2022} demonstrated the extent to which hydrogen and electricity grid expansion is interchangeable in a climate-neutral European energy system at the expense of higher investments. The underlying, highly-resolved model encompasses the most relevant sectors, however it neglects the infrastructure required to transport carbon accross regions.
Morbee et al.~\cite{morbeeOptimisedDeploymentEuropean2012} optimizes the topology and capacity of a \carbon{} network in Europe, but only considers the power sector without co-optimizing renewable deployment. This limited sectoral scope cannot capture important dynamics of carbon management, since it neglects sectors like industry that will need to handle most \carbon{} in the future.
The European Commission's Joint Research Center analysed the possible topology of a \carbon{} network in 2024~\cite{jrc2024}, but did not include the interaction of  \carbon{} sources and sinks with the rest of the energy system.
Becattini et al~\cite{becattiniCarbonDioxideCapture2022} present a mixed-integer model to optimize the time-evolution of a \carbon{} transport system in Switzerland, connected to a remote sequestration site in Norway. However, this limited spatial scope fails to consider other sequestration sites and the co-benefits from connecting the \carbon{} network to neighboring countries.
Other high-resolution energy system models do not feature detailed CU technologies and \carbon{} transport. In Pickering et al.~\cite{pickeringDiversityOptionsEliminate2022}, the Euro-Calliope model, for example, features neither carbon nor hydrogen transport, and thus forces generic synthetic fuel production to rely on locale electrolysis and DAC in the same region.

No study has yet considered the co-optimization and comprehensive assessment of both \carbon{} and \hydrogen{} networks in a fully sector-coupled energy system. Such an assessment is strongly needed to identify realms of competition and synergy between hydrogen and carbon management technologies. In this paper, we present a detailed study of the European energy system for 2050, which includes high geographical resolution and a comprehensive representation of carbon management technologies. The study is conducted using the PyPSA-Eur energy system model~\cite{brownPyPSAEurOpenSectorcoupled2023} which optimizes operations and investments in all relevant energy sectors to supply projected energy demands for 2050 (see Fig.~\ref{fig:total-demand-bar}). It is assumed that Europe is self-sufficient in energy and does not import any fuels. This drives carbonaceous fuel production to be located at sites where both \carbon{} and \hydrogen{} can be provided at low cost.

@@

We analyze how the option to deploy carbon and hydrogen networks affects different decarbonization strategies in the scenarios and how investments in none or only one of the transport technologies lead to inefficient infrastructure layouts. Initially, we focus on a net-zero carbon emissions target, limiting carbon sequestration to 200~Mt/a. This constraint ensures that carbon sequestration is reserved for the hardest-to-abate emissions, while avoiding reliance on sequestration where other mitigation options are feasible. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened target with net \carbon{} removal.

@@

All scenarios share key characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 764~bn€/a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.7\% in the \hydrogengrid{}, and 5.3\% in the \hybridscenario{}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\linewidth]{comparison/default/figures/90_nodes/cost_bar.png}
    \caption[short]{Total annual system cost, subdivided into groups of technologies for the different models of the European energy system with a net-zero emission target. While in the \baselinescenario{}, the model has neither a carbon nor a hydrogen network, it can expand both in the \hybridscenario{}. ``Gas Infrastructure'' combines gas facilities for transport, power and heat production, ``\carbon{} Infrastructure'' and ``\hydrogen{} Infrastructure'' combine transport and storage for each carrier. ``Carbon Capture at Point Sources'' combines all technologies with integrated carbon capture, including the cost of the main facility (e.g., CHP units) and the carbon capture application.}
    \label{fig:cost_bar}
\end{figure}

@@

System cost differences between the scenarios are driven by investments in wind and solar, as well as carbon capture technologies. In scenarios with carbon and hydrogen networks, the model primarily reduces the reliance on costly DAC, from 365~Mt/a carbon capture in the \baselinescenario{} to 113~Mt/a in the \hybridscenario{} (see Fig.~\ref{fig:balance_captured_carbon}). At the same time, carbon capture from biomass combustion, gas-based industry processes and biogas upgrading increase. The latter lowers the need for synthetic methane production.
% Another knock-on effect of less reliance on DAC is its reduced endogenous demand for dedicated renewables as well as heat pumps and biomass CHPs to supply power and heat.

@@

In the \baselinescenario{}, where there is no carbon and hydrogen transport, hydrogen must be produced where it is consumed. For CU, either hydrogen must be produced where carbon is cheap at point sources, or DAC must be used at sites with low-cost hydrogen. Where fossil carbon from point sources cannot be sequestered, it has to be compensated elsewhere by negative emissions from DAC or biomass capture with sequestration. To provide hydrogen in some regions with poor renewable resources the model resorts to SMR with either green or fossil methane transported in the gas grid.All of these options are sub-optimal from a cost perspective.

@@

In the \carbonscenario, there are three main transport purposes of the low-cost carbon captured in Central and Eastern Europe (see Fig.~\ref{fig:balance_map}, top left). A large part of the carbon is transported to the Iberian Peninsula (avg. 9.8~kt/h) and the British Isles (avg. 11.8~kt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. In Italy and Greece, a portion of the carbon (avg. 3.3~kt/h) is directed to methanation facilities where the synthetic methane is fed into the gas grid at key terminals that also receive fossil gas imports. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. Almost none of the captured carbon is used locally at the point source. In addition to electrolysis in regions with good renewable resources to supply CU ($\sim$125~GW$_\text{el}$), the model places smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$8~GW$_\text{el}$) (see bottom left). Hydrogen prices vary widely across the continent, with low price ``valleys'' at around 60~€/MWh, where the CU is located, and high price ``peaks'' in Central Europe at around 95~€/MWh. Despite the transport system, marginal cost of carbon capture still vary from 105~€/t in Central Europe to 160~€/t at sequestration sites.

In the \hydrogenscenario{}, the main hydrogen transportation route goes from Spain (66~GW) to Central Europe to supply local carbonaceous fuel production with captured carbon from point sources, since \carbon{} can no longer be transported (see Fig~\ref{fig:balance_map}, bottom right). Also, the hydrogen network supplies spatially fixed hydrogen demand for industry in Central Europe (2.5~GW in total) at much lower costs than local electrolysis or SMR. Carbon sequestration takes place at several coastal sites with local DAC facilities. The largest carbon sink is located in Portugal (3.5~kt/h, see Fig~\ref{fig:balance_map}, top right). Marginal costs of carbon capture are low in Central Europe (110-120~€/t) and highest in sequestration regions (160~€/t). Hydrogen prices have flattened out at a relatively low level of €60~/MWh in all regions.
